Accelerating Biomolecular Nuclear Magnetic Resonance Assignment with A*

### 2014-03-27 paper/tex/Paper.tex
paper
Paper Revision

\section{Introduction}
\label{sub:introduction}
Knowledge of a protein’s structure allows for further understanding of the function of a protein and alterations to its function that can lead to disease. The post-genomic era has brought the need to gather functional and structural information of unknown proteins that are encoded by newly discovered genes. Through the implementation of Nuclear magnetic resonance (NMR) spectroscopy, the structure of molecules, such as proteins, can be studied. However, NMR spectroscopy generates large amounts of data that needs to be assigned. The assignment process can take months to complete and is prone to error. To allow for major advances in structural biology, a faster and more accurate method needs to be developed. Our research focuses on developing an algorithm to automate the assignment process, in an attempt to minimize human error and accelerate the time consuming task.

\subsection{NMR} % (fold)
\label{sub:nmr}
Several types of NMR variables can be used in the analysis of protein structures. In particular, essential information is provided by the chemical shifts of NMR-active nuclei present in proteins, including hydrogen and isotopes of carbon and nitrogen. The chemical shift is a quantifier for the deviation in the resonant frequency of a nucleus from its value in a structure-free environment, and therefore provides information on the local conformation. Determining the chemical shifts of all or most of the nuclei in a biomolecule is the first step in determining its structure. An important set of chemical shifts in a protein are those corresponding to the nuclei in the backbone of the protein polypeptide chain, including the nitrogen, attached hydrogen, and the alpha and beta carbon atoms ($C_\alpha$ and $C_{\beta}$) of each residue. These chemical shifts are measured using various three-dimensional NMR experiments, and then matched to the individual residues in the protein in a process called sequential assignment.

@@

\section{Method} % (fold)
\label{sec:method}

\subsection{Algorithm} % (fold)
\label{sub:algorithm}
Our algorithm begins by reading in an expected amino acid order and the NMR carbon shift data from a data file. It then changes the amino acids to their expected $C_\alpha$ and $C_{\beta}$ values and stores them in a list that represents the protein chain. Next, the carbon shift values for a single amino acid are placed in a tile. A tile contains the $C_\alpha$ and $C_{\beta}$ values for the amino acid residue $i$ and residue $i-1$. Next, missing data is accounted for. If there are fewer tiles than the number of amino acids in the protein chain, placeholder tiles are generated to make up the difference. Before the search for the best solution begins, the tiles are grouped by the types of amino acids they could represent. The groups were created using the expected $C_\alpha$ and $C_{\beta}$ values shown in Figure~\ref{fig:carbon}. Grouping is done in an attempted to accelerate the assignment process. Tiles that do not match a group are grouped with the placeholder tiles.

\begin{figure}[H]
\begin{center}
\includegraphics{carbon}
\end{center}
\caption{Expected Carbon Shift Values}
\label{fig:carbon}
\end{figure}

@@

\subsection{Results}
\label{sub:results}
Our algorithm has been tested with real NMR datasets of varying size. When compared to hand-ordered data, the algorithm produces the same, optimal ordering of amino acids. Initially, we focused our research on accurate assignment at the cost of speed. Thus, at the start of our research we could assign only smaller datasets up to ten to fifteen amino acids long. As we shifted our focus to optimizing the algorithm's speed while maintaining the accuracy we managed to achieve we improved runtime to correctly assign up to forty to fifty amino acids. In future research, we plan to test with larger datasets, maximizing speed and efficiency to the point that our algorithm can assign full protein sequences. We also plan to look into testing datasets with more than a few missing datapoints to verify that our algorithm correctly places amino acids when given less-perfect data.

\subsection{Future Goals} % (fold)
\label{sub:future_goals}
We are currently looking into the possibility of parallelizing our algorithm. If possible, our algorithm will be able to handle full NMR datasets. We are also looking to use machine learning in order to optimize the cost calculation. 
% subsection future_goals (end)

>>>>

\section{Introduction}
\label{sub:introduction}
Knowledge of a protein’s structure allows for further understanding of the function of a protein and alterations to its function that can lead to disease. Through the implementation of Nuclear magnetic resonance (NMR) spectroscopy, the structure of molecules, such as proteins, can be studied. However, assigning NMR spectroscopy generates large amounts of data that needs to be assigned. This process can take months to complete and is prone to error. To allow for major advances in structural biology, a faster and more accurate method needs to be developed. Our research focuses on developing an algorithm to automate the assignment process. We hope our attempts can minimize the human error and decrease the time consumption associated with manual assignment , increasing the overall effectiveness of the operation.

\subsection{Nuclear Magnetic Resonance Spectroscopy} % (fold)
\label{sub:nmr}
Several types of NMR variables can be used in the analysis of protein structures. In particular, essential information is provided by the chemical shifts of NMR-active nuclei present in proteins, including hydrogen and isotopes of carbon and nitrogen. The chemical shift is a quantifier for the deviation in the resonant frequency of a nucleus from its value in a structure-free environment, and therefore provides information on the local conformation. Determining the chemical shifts of all or most of the nuclei in a biomolecule is the first step in determining its structure. An important set of chemical shifts in a protein are those corresponding to the nuclei in the backbone of the protein polypeptide chain, including the nitrogen, attached hydrogen, and the alpha and beta carbon atoms ($C_\alpha$ and $C_{\beta}$) of each residue. These chemical shifts are measured using various three-dimensional NMR experiments, and then matched to the individual residues in the protein in a process called sequential assignment.

@@

\section{Assignment Methods} % (fold)
\label{sec:method}

\subsection{Manual Assignment} % (fold)
\label{sub:manual_assignment}
% TODO
add this part
cite our sources

% subsection manual_assignment (end)

\subsection{Our Algorithm} % (fold)
\label{sub:algorithm}
Our algorithm begins by reading in an expected amino acid order and the NMR carbon shift data from a data file. It then changes the amino acids to their expected $C_\alpha$ and $C_{\beta}$ values and stores them in a list that represents the protein chain. Next, the carbon shift values for a single amino acid are placed in a tile. A tile contains the $C_\alpha$ and $C_{\beta}$ values for the amino acid residue $i$ and residue $i-1$. Next, missing data is accounted for. If there are fewer tiles than the number of amino acids in the protein chain, placeholder tiles are generated to make up the difference. Before the search for the best solution begins, the tiles are grouped by the types of amino acids they could represent. The groups were created using the expected $C_\alpha$ and $C_{\beta}$ values shown in Figure~\ref{fig:carbon}. Grouping is done in an attempted to accelerate the assignment process. Tiles that do not match a group are grouped with the placeholder tiles.

\begin{figure}[H]
\begin{center}
\includegraphics{carbon}
\end{center}
\caption{Expected Carbon Shift Values} % TODO citation
\label{fig:carbon}
\end{figure}

@@

\subsection{Results}
\label{sub:results}
Our algorithm has been tested with NMR datasets of varying size from experiments conducted by Dr. Kilpatrick.  When compared to manually assigned data, the algorithm produces the same, optimal ordering of amino acids. Initially, we focused our research on accurate assignment at the cost of speed. Thus, at the start of our research our algorithm was only able to assign smaller datasets with a chain of ten to fifteen amino acids. Shifting our focus to optimizing the algorithm's speed while maintaining the accuracy we had managed to achieve, we were able to improve runtime. With the increased runtime, we were able to correctly assign forty to fifty amino acids. In future research, we plan to test with larger datasets, maximizing speed and efficiency to the point that our algorithm can assign full protein(amino acid?) sequences. We also plan to look into testing datasets with more than a few missing data points to verify that our algorithm correctly places amino acids when given less perfect data.  - TIME FOR DATASETS?

\subsection{Future Goals} % (fold)
\label{sub:future_goals}
In future research, we will look into maximizing speed and efficiency to the point that our algorithm can assign full amino acid sequences. In order to achieve this, we will experiment with  both parallelization (techniques) and machine learning. The latter of these two will assist in optimizing the cost calculation while the former will increase the speed of assignment. We believe these implementations will help our algorithm handle datasets with numerous missing data points, validating our results even when given incomplete data. 
% subsection future_goals (end)
